Compiling refinement types to dependent types

### 2020-07-22 tex/paper.tex
Move the "in words" part before the formal definition of tranty
Remove the attempt at extending Boehm-Berarducci to dependent pattern-matching

We also establish some derived forms in the core language to simplify the subsequent exposition,
and state some straightforward facts that will be useful later:
\begin{itemize}
  \item Simplified variable typing rule which we call \textsc{CT-VarW}:
    \[
      \inferrule{\nonterm{GC} \vdash \nonterm{ec} : s \\ \gterm x : \nonterm{ec} \in \nonterm{GC}}
                {\nonterm{GC} \vdash \gterm x : \nonterm{ec}}.
    \]
    This rule unfolds into a single \textsc{CT-Var} defining $x$,
    followed by a sequence of \textsc{CT-Weaken}
    to ``move'' the $x$ into the right position in the final $\nonterm{GC}$.
  \item Non-dependent function type:
    \[
      \nonterm{tc}_1 \rightarrow \nonterm{tc}_2 \triangleq \Pi \_ : \nonterm{tc}_1. \nonterm{tc}_2.
    \]
  \item Dependent pair type, or $\Sigma$-type, with the first component being a value $x$ of type $\tau$ and the second component being of type $P x$:
    \[
      (x : \nonterm{tc} \dast P x) \triangleq \Pi \alpha : \star. (\Pi x : \nonterm{tc}. P x \rightarrow \alpha) \rightarrow \alpha,
    \]
    with the corresponding constructor (we overload the syntax here to resemble Idris):
    \[
      (x \dast p) \triangleq \lambda \alpha : \star. \lambda f : (\Pi x' : \nonterm{tc}. P x' \rightarrow \alpha). f x p
    \]
    and projections
    \[
      \text{fst} \triangleq \lambda \sigma : (x : \nonterm{tc} \dast P x). \sigma \nonterm{tc} (\lambda x' : \nonterm{tc}. \lambda \_ : P x'. x'),
    \]
    \[
      \text{snd} \triangleq \lambda \sigma : (x : \nonterm{tc} \dast P x). \sigma \nonterm{tc} (\lambda x' : \nonterm{tc}. \lambda p : P x'. p).
    \]
    It's easy to see from the formation rule that
    \begin{lemma}
      If $\nonterm{tc}$ and $P x$ are well-formed in $\nonterm{GC}$,
      then $(x : \nonterm{tc} \dast P x)$ is well-formed in $\nonterm{GC}$ and has type $\star$.
    \end{lemma}
  \item Non-dependent pair type:
    \[
      (\nonterm{tc}_1, \nonterm{tc}_2) \triangleq \{ x : \nonterm{tc}_1\ **\ (\lambda \_ : \nonterm{tc}_1. \nonterm{tc}_2) x \}.
    \]
    As a direct consequence of the previous lemma we obtain
    \begin{lemma}\label{lma:non_dep_pair_typing}
      If $\nonterm{tc}_1, \nonterm{tc}_2$ are well-formed in $\nonterm{GC}$,
      then $(\nonterm{tc}_1, \nonterm{tc}_2)$ is well-formed in $\nonterm{GC}$ and has type $\star$.
    \end{lemma}
  \item Equality of expressions $\nonterm{ec}_1, \nonterm{ec}_2$ of type $S$:
    \[
      \nonterm{ec}_1 \equiv \nonterm{ec}_2 \triangleq \Pi P : S \rightarrow \star. (P x \rightarrow P y, P y \rightarrow P x).
    \]
    The reader might recognize this as the Leibniz equality \cite{FindSomethingForThis}.
  \item ADTs, expressed using an encoding similar to Boehm-Berarducci's \cite{Bohm85}.
    We borrow the syntax for the ADTs from the surface language and put a hat on it,
    so let's take some ADT $\widehat{\langle \overline{l_i : \nonterm{ts}_i}^i \rangle}$.
    The usual Boehm-Berarducci encoding associates the type
    $\Pi \alpha : \star. \overline{(\nonterm{ts}_i \rightarrow \alpha)}^i \rightarrow \alpha$ with it.

    We go a step further to account for the subset of the dependent pattern matching
    that we are interested in.

    First, each label $l_i$ gives rise to a term $\hat{l_i}$ of type $\nonterm{ts}_i \rightarrow \xi$.
    %TODO operational semantics?
    We then define the derived form
    \[
      \widehat{\langle \overline{l_i : \nonterm{ts}_i}^i \rangle}
        \triangleq
          \Pi s : \xi.
          \Pi \alpha : \star.
          \overline{(\Pi x_i : \nonterm{ts}_i. s \equiv \hat{l_i} x_i \rightarrow \alpha)}^i \rightarrow \alpha.
    \]
    The outer extra argument named $s$ denotes the scrutinee of the case expression to be defined next.
    The inner extra argument with the type $s \equiv \hat{l_i} x_i$
    expresses the subset of dependent pattern matching
    that we rely on for the translation from the surface language:
    in particular, it provides the ``case branch'' with a proof
    that the scrutinee $s$ is indeed equal to the ``case pattern'' $\hat{l_i} x_i$.
\end{itemize}

@@

\paragraph{Types.}
Next, we define $\Tranty$.
We now consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$.

@@

Or, in words:
\begin{itemize}
  \item (dependent) arrow types are translated to the corresponding $\Pi$-types,
  \item refined types are translated to the corresponding $\Sigma$-types,
  \item atomic refinements, being propositions about surface language terms equality,
    are translated to propositions stating the equality of translated core language terms,
  \item conjunctions of several (atomic) refinements are represented as tuples of the corresponding atomic translations.
\end{itemize}

>>>>

We also establish some derived forms in the core language to simplify the subsequent exposition,
and state some straightforward facts that will be useful later:
\begin{itemize}
  \item Simplified variable typing rule which we call \textsc{CT-VarW}:
    \[
      \inferrule{\nonterm{GC} \vdash \nonterm{ec} : s \\ \gterm x : \nonterm{ec} \in \nonterm{GC}}
                {\nonterm{GC} \vdash \gterm x : \nonterm{ec}}.
    \]
    This rule unfolds into a single \textsc{CT-Var} defining $x$,
    followed by a sequence of \textsc{CT-Weaken}
    to ``move'' the $x$ into the right position in the final $\nonterm{GC}$.
  \item Non-dependent function type:
    \[
      \nonterm{tc}_1 \rightarrow \nonterm{tc}_2 \triangleq \Pi \_ : \nonterm{tc}_1. \nonterm{tc}_2.
    \]
  \item Dependent pair type, or $\Sigma$-type, with the first component being a value $x$ of type $\tau$ and the second component being of type $P x$:
    \[
      (x : \nonterm{tc} \dast P x) \triangleq \Pi \alpha : \star. (\Pi x : \nonterm{tc}. P x \rightarrow \alpha) \rightarrow \alpha,
    \]
    with the corresponding constructor (we overload the syntax here to resemble Idris):
    \[
      (x \dast p) \triangleq \lambda \alpha : \star. \lambda f : (\Pi x' : \nonterm{tc}. P x' \rightarrow \alpha). f x p
    \]
    and projections
    \[
      \text{fst} \triangleq \lambda \sigma : (x : \nonterm{tc} \dast P x). \sigma \nonterm{tc} (\lambda x' : \nonterm{tc}. \lambda \_ : P x'. x'),
    \]
    \[
      \text{snd} \triangleq \lambda \sigma : (x : \nonterm{tc} \dast P x). \sigma \nonterm{tc} (\lambda x' : \nonterm{tc}. \lambda p : P x'. p).
    \]
    It's easy to see from the formation rule that
    \begin{lemma}
      If $\nonterm{tc}$ and $P x$ are well-formed in $\nonterm{GC}$,
      then $(x : \nonterm{tc} \dast P x)$ is well-formed in $\nonterm{GC}$ and has type $\star$.
    \end{lemma}
  \item Non-dependent pair type:
    \[
      (\nonterm{tc}_1, \nonterm{tc}_2) \triangleq \{ x : \nonterm{tc}_1\ **\ (\lambda \_ : \nonterm{tc}_1. \nonterm{tc}_2) x \}.
    \]
    As a direct consequence of the previous lemma we obtain
    \begin{lemma}\label{lma:non_dep_pair_typing}
      If $\nonterm{tc}_1, \nonterm{tc}_2$ are well-formed in $\nonterm{GC}$,
      then $(\nonterm{tc}_1, \nonterm{tc}_2)$ is well-formed in $\nonterm{GC}$ and has type $\star$.
    \end{lemma}
  \item Equality of expressions $\nonterm{ec}_1, \nonterm{ec}_2$ of type $S$:
    \[
      \nonterm{ec}_1 \equiv \nonterm{ec}_2 \triangleq \Pi P : S \rightarrow \star. (P x \rightarrow P y, P y \rightarrow P x).
    \]
    The reader might recognize this as the Leibniz equality \cite{FindSomethingForThis}.
\end{itemize}

@@

\paragraph{Types.}
Next, we define $\Tranty$. In short:
\begin{itemize}
  \item (dependent) arrow types are translated to the corresponding $\Pi$-types,
  \item refined types are translated to the corresponding $\Sigma$-types,
  \item atomic refinements, being propositions about surface language terms equality,
    are translated to propositions stating the equality of translated core language terms,
  \item conjunctions of several (atomic) refinements are represented as tuples of the corresponding atomic translations.
\end{itemize}

More formally, we consider the last rule used in the derivation of $\nonterm G \vdash \nonterm{ts}$.

@@
